Deterministic and statistical truncated KdV models for anomalous waves induced by abrupt depth change

### 2018-09-25 LatexNotes/Notes.tex
NM
Typing up notes on non-dimensionalization.

% Experimental parameters
\newcommand{\lamm}{\lambda^{-}}
\newcommand{\lamp}{\lambda^{+}}
\newcommand{\hm}{h^{-}}
\newcommand{\hp}{h^{+}}

% Theory for truncated KdV and stat mech
\newcommand{\uhat}{\hat{u}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\Real}{\text{Re}}
\newcommand{\Fspace}{\mathscr{F}_{\Lambda}}
\newcommand{\Proj}{P_{\Lambda}}
\newcommand{\sumk}{\sum_{k=1}^{\Lambda}}
\newcommand{\sumn}{\sum_{n=1}^{N}}
\newcommand{\uDir}{u_{\text{Dir}}}
\newcommand{\Gibbs}{\mathcal{G}}

@@

Following Johnson's derivation the dimensionless KdV equation is (on page 208)
\begin{equation}
2 \eta_{\tau} + 3 \eta \eta_{\xi} + \frac{1}{3} \eta_{\xi \xi \xi} = 0
\end{equation}
I had to work backwards quite a bit to figure out how this was non-dimensionalized (Johnson keeps rescaling things over and over again). Ultimately, the characteristic scales are
\begin{align}
& Z^* = h_0 , \quad X^* = \eps^{-1/2} h_0 , \quad T^* = \eps^{-1/2} \sqrt{h_0/g} , \\
& W^* = \eps^{3/2} \sqrt{g h_0} , \quad U^* = \eps \sqrt{g h_0} , \quad P^* = \eps \rho g h_0
\end{align}
where $\eps$ is the ratio of characteristic amplitude to depth
\begin{equation}
\eps = a / h_0
\end{equation}
The variables $\tau$ and $\xi$ are
\begin{align}
\tau = \eps t , \quad \xi = x-t
\end{align}
Remember, the privileged KdV scaling is $\delta^2 \sim \eps$ where $\delta = h_0 / \lambda$. If this scaling is inserted into above, then $X^* \sim \lambda$, $T^* \sim \lambda/\sqrt{g h_0}$ which all makes sense.

>>>>

% Experimental parameters
\newcommand{\lamm}{\lambda^{-}}
\newcommand{\lamp}{\lambda^{+}}
\newcommand{\hm}{h^{-}}
\newcommand{\hp}{h^{+}}
\newcommand{\etastd}{\eta_{\text std}}
% Theory for truncated KdV and stat mech
\newcommand{\uhat}{\hat{u}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\Real}{\text{Re}}
\newcommand{\Fspace}{\mathscr{F}_{\Lambda}}
\newcommand{\Proj}{P_{\Lambda}}
\newcommand{\sumk}{\sum_{k=1}^{\Lambda}}
\newcommand{\sumn}{\sum_{n=1}^{N}}
\newcommand{\uDir}{u_{\text{Dir}}}
\newcommand{\Gibbs}{\mathcal{G}}

@@

\subsubsection{Johnson's formulation}
Following Johnson's derivation the dimensionless KdV equation is (on page 208)
\begin{equation}
2 \eta_{\tau} + 3 \eta \eta_{\xi} + \frac{1}{3} \eta_{\xi \xi \xi} = 0
\end{equation}
I had to work backwards quite a bit to figure out how this was non-dimensionalized (Johnson keeps rescaling things over and over again). Ultimately, the characteristic scales are
\begin{align}
& Z^* = h_0 , \quad X^* = \eps^{-1/2} h_0 , \quad T^* = \eps^{-1/2} \sqrt{h_0/g} , \\
& W^* = \eps^{3/2} \sqrt{g h_0} , \quad U^* = \eps \sqrt{g h_0} , \quad P^* = \eps \rho g h_0
\end{align}
where $\eps$ is the ratio of characteristic amplitude to depth
\begin{equation}
\eps = a / h_0
\end{equation}
The variables $\tau$ and $\xi$ are
\begin{align}
\tau = \eps t , \quad \xi = x-t
\end{align}
Remember, the privileged KdV scaling is $\delta^2 \sim \eps$ where $\delta = h_0 / \lambda$. If this scaling is inserted into above, then $X^* \sim \lambda$, $T^* \sim \lambda/\sqrt{g h_0}$ which all makes sense.

If we convert KdV back into dimensional form, we get
\begin{equation}
2 \eta_t + \frac{3 c}{h_0} \eta \eta_{\xi} + \frac{c h_0^2}{3} \eta_{\xi \xi \xi} = 0
\end{equation}
Here, $\eta$ has units of length, $t$ time, $c = \sqrt{g h_0}$ is the wave speed, and $\xi = x - ct$.

\subsubsection{My own reformulation}

We now seek to non-dimensionalize the KdV equation, but not exactly the same way that Johnson did. Some things we would like to accomplish are:
\begin{enumerate}
\item When recast in dimensionless variables, $\eta$ should be periodic on the interval $[0, 2\pi)$ in order to make the Fourier notation simple.
\item We want to rescale $\eta$ to have unit energy in order to make the Hamiltonian sampling simple.
\end{enumerate}
First, we observe there are a few lengthscales in the experiment that may be relevant: the upstream depth $h^- = 12.5$ cm, the tank width $L_0 = 20$ cm, the peak wavelength $\lambda_p^- = 38$ cm (induced by the peak forcing frequency as calculated by the dispersion relation), and possibly the maximum wavelength present $\lambda_{max}^{-} = ?$. First, since we want $\eta$ to be periodic on $[0,2 \pi)$ in dimensionless variables, the most logical choice for scaling $x$ is a wavelength, either the peak or max. Lets choose the peak. Next, since we want dimensionless $\eta$ to have unit energy, we {\em must} scale $\eta$ by the characteristic amplitude $a = \etastd = \mean{\eta^2}^{1/2}$, where the mean is taken over a sufficiently long time or distance. We therefore introduce dimensionless variables, denoted by a tilde
\begin{align}
& x = \frac{\lambda_p}{2\pi} \tilde{x} , \quad 
t = \frac{\lambda_p}{2\pi c} \tilde{t} , \quad
\eta = a \tilde{\eta} \\
& \xi = x-ct = \frac{\lambda_p}{2\pi} (\tilde{x} - \tilde{t}) = \frac{\lambda_p}{2\pi} \tilde{\xi}
\end{align}

Then in dimensionless variables (after dropping the tildes), the system is given by
\begin{equation}
\eta_t + \frac{3}{2}\eps \eta \eta_{\xi} + \frac{2 \pi^2}{3} \delta^2 \eta_{\xi \xi \xi} = 0
\end{equation}
where we have the dimensionless variables
\begin{align}
\eps = a/h_0 , \quad \delta = h_0/\lambda_p
\end{align}
In these dimensionless variables, we have that $\eta$ is periodic for $x \in [0, 2\pi)$ and we have $\mean{\eta^2} = 1$, just as intended.
Note, in the experiments, $\delta = 0.33$ and $\eps$ lies in the range $2.4 \times 10^{-3}$ to $2.4 \times 10^{-2}$.

We will eventually want to rescale $t$, to look at long-time scales as is typically done in KdV analysis. Accordingly, we will define
\begin{equation}
\tau = \beta t
\end{equation}
where $\beta$ is usually either $\eps$ or $\delta^2$ (or we can choose it to be something more appropriate for the experiments). Of course, the effect of this rescaling would be to introduce a factor of $\beta$ in front of the first term $\eta_{\tau}$.

Finally, since we want to study variable-depth KdV, we will ultimately want to interchange the roles of space and time as Johnson does on p. 269. So we will have to modify some of this non-dimensionalization. I had to wrap my mind around the classical case first though.

\subsubsection{Comparison to variables in Andy's original notes}
In Andy's original notes (as well as in Di Qi's latex notes), the width of the tank $L_0$ was used for non-dimensionalizing $x$ (as well as $\xi$). In comparing our two formulations, the main subtlety is the energy, defined as
\begin{equation}
E_0 = \frac{1}{2} \int_{0}^{2 \pi L_0} u^2 dx
\end{equation}
where $u$ is displacement and has units of length. Thus, $E_0$ has units of length cubed (it took me a while to realize this is the main subtlety in reconciling the two formulations). The standard deviation of displacement is
\begin{equation}
u_{std} = \frac{1}{\sqrt{2 \pi L_0}} \left( \int_{0}^{2\pi L_0} u^2 dx \right)^{1/2}
\end{equation}
or in Andy's / Di Qi's notation
\begin{equation}
u_{std} = const \times L_0^{-1/2} E_0^{1/2}
\end{equation}
where I really don't care what the constant is. The main thing I wanted to see was why $L_0^{-1/2}$ appears, and the above shows this.

The main difficulty I have with this formulation is that there is no reason to expect the spatial period of $u$ to have anything to do with the tank width $L_0$. More logically, the spatial period of $u$ should correspond to a characteristic wavelength in the problem, either the peak or maximum wavelength. The latter is more likely to be true in a literal sense, but I believe the peak wavelength provides a more reasonable scale for non-dimensionalization.
